---
title: "US Storm Data"
author: "Hussam Zarea"
date: "July 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#set directory
```

setwd("C:/Users/hussamzarea/Desktop/Coursera/ReproducibleResearch/Week4")
```

```{r}
#download data file
#download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "stormData.csv.bz2")
stormDatra <- read.csv(bzfile("stormData.csv.bz2"), sep=",", header=T)
```

```{r}
#get data diminition
dim(stormDatra)
```

```{r}
#get data header
head(stormDatra)
```

```{r}
#get data needed
cleanStormData <- stormDatra[,c(8,23:28)]
```

```{r}
# View headers in revised data set
head(cleanStormData)
```

```{r}
# Convert Property Damage 
cleanStormData$PROPDMGDOLLARS = 0
cleanStormData[cleanStormData$PROPDMGEXP == "H", ]$PROPDMGDOLLARS = cleanStormData[cleanStormData$PROPDMGEXP == "H", ]$PROPDMG * 10^2
cleanStormData[cleanStormData$PROPDMGEXP == "K", ]$PROPDMGDOLLARS = cleanStormData[cleanStormData$PROPDMGEXP == "K", ]$PROPDMG * 10^3
cleanStormData[cleanStormData$PROPDMGEXP == "M", ]$PROPDMGDOLLARS = cleanStormData[cleanStormData$PROPDMGEXP == "M", ]$PROPDMG * 10^6
cleanStormData[cleanStormData$PROPDMGEXP == "B", ]$PROPDMGDOLLARS = cleanStormData[cleanStormData$PROPDMGEXP == "B", ]$PROPDMG * 10^9

# Convert Crop Damage 
cleanStormData$CROPDMGDOLLARS = 0
cleanStormData[cleanStormData$CROPDMGEXP == "H", ]$CROPDMGDOLLARS = cleanStormData[cleanStormData$CROPDMGEXP == "H", ]$CROPDMG * 10^2
cleanStormData[cleanStormData$CROPDMGEXP == "K", ]$CROPDMGDOLLARS = cleanStormData[cleanStormData$CROPDMGEXP == "K", ]$CROPDMG * 10^3
cleanStormData[cleanStormData$CROPDMGEXP == "M", ]$CROPDMGDOLLARS = cleanStormData[cleanStormData$CROPDMGEXP == "M", ]$CROPDMG * 10^6
cleanStormData[cleanStormData$CROPDMGEXP == "B", ]$CROPDMGDOLLARS = cleanStormData[cleanStormData$CROPDMGEXP == "B", ]$CROPDMG * 10^9

```

```{r}
# get data headers to check the change
head(cleanStormData)
```

```{r}
#load libraries
library(ggplot2)
library(gridExtra)
```

```{r}
#group fatalities by event type and sum it
fatalities <- aggregate(FATALITIES ~ EVTYPE, data=cleanStormData, sum)
```

```{r}
#group injuries by event type and sum it
injuries <- aggregate(INJURIES ~ EVTYPE, data = cleanStormData, sum)
```

```{r}
## sort the fatality data
fatalities <- fatalities[order(-fatalities$FATALITIES), ][1:20, ]
fatalities$EVTYPE <- factor(fatalities$EVTYPE, levels = fatalities$EVTYPE)
head(fatalities)
```

```{r}
## Sort the injury data
injuries <- injuries[order(-injuries$INJURIES), ][1:20, ]
injuries$EVTYPE <- factor(injuries$EVTYPE, levels = injuries$EVTYPE)
head(injuries)
```


```{r}
# set the fatality plot pramaters
fatalityPlot = ggplot(fatalities, aes(x = EVTYPE, y = FATALITIES, theme_set(theme_bw()))) + 
  geom_bar(stat = "identity", fill = "red") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 4)) + 
  xlab("Event Type") + 
  ylab("Fatalities") + 
  ggtitle("Fatality by Event Types") +
  theme(plot.title = element_text(size = 10))
```

```{r}
# set the injury plot pramaters
injuryPlot = ggplot(injuries, aes(x = EVTYPE, y = INJURIES, theme_set(theme_bw()))) + 
  geom_bar(stat = "identity", fill = "blue") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 4)) + 
  xlab("Event Type") + 
  ylab("Injuries") + 
  ggtitle("Injury by Event Types") +
  theme(plot.title = element_text(size = 10))
```

```{r}
# plot both fatality and injury side by side
grid.arrange(fatalityPlot, injuryPlot, ncol = 2, top = "Most Harmful Events By Population Health")
```

```{r}
## Organize Property & Crop to Event Type and store in object "damage"
damage <- aggregate(PROPDMGDOLLARS + CROPDMGDOLLARS ~ EVTYPE, data=cleanStormData, sum)
names(damage) = c("EVENT_TYPE", "TOTAL_DAMAGE")
```

```{r}
#get the most damage event in the US 
damage <- damage[order(-damage$TOTAL_DAMAGE), ][1:20, ]
damage$EVENT_TYPE <- factor(damage$EVENT_TYPE, levels = damage$EVENT_TYPE)
```

```{r}
# get header
head(damage)
```

```{r}
#plot the most damage event in the US
ggplot(damage, aes(x = EVENT_TYPE, y = TOTAL_DAMAGE, theme_set(theme_bw()))) +
  geom_bar(stat = "identity", fill = "red") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  xlab("Event Type") + 
  ylab("Total Damage in US Dollor") + 
  ggtitle("the US greatest economic consequences in Property & Crop Damage by top 20 Weather Events")
```