---
title: "US Storm Data"
author: "Hussam Zarea"
date: "July 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#Reproducible Research Week 4

##Data Processing

####Set The Directory
```{r}
setwd("C:/Users/hzarea/Desktop/Coursera/ReproducibleResearch/Week4")
```

####Download and read the data and store it in stormData variable 
```{r}
#download data file
#use this one time only.  after the data is downloaded, coment it
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", destfile = "stormData.csv.bz2")

#read data and store in stormData
stormDatra <- read.csv(bzfile("stormData.csv.bz2"), sep=",", header=T)
```

####Get the ddiminition of the stormData
```{r}
dim(stormDatra)
```

####Get the header of the stormData
```{r}
head(stormDatra)
```

####Get only the needed stormData
```{r}
cleanStormData <- stormDatra[,c(8,23:28)]
```

####View the headers of the cleanStormData
```{r}
head(cleanStormData)
```

####Convert the property damage
```{r}
cleanStormData$PROPDMGDOLLARS = 0
cleanStormData[cleanStormData$PROPDMGEXP == "H", ]$PROPDMGDOLLARS = cleanStormData[cleanStormData$PROPDMGEXP == "H", ]$PROPDMG * 10^2
cleanStormData[cleanStormData$PROPDMGEXP == "K", ]$PROPDMGDOLLARS = cleanStormData[cleanStormData$PROPDMGEXP == "K", ]$PROPDMG * 10^3
cleanStormData[cleanStormData$PROPDMGEXP == "M", ]$PROPDMGDOLLARS = cleanStormData[cleanStormData$PROPDMGEXP == "M", ]$PROPDMG * 10^6
cleanStormData[cleanStormData$PROPDMGEXP == "B", ]$PROPDMGDOLLARS = cleanStormData[cleanStormData$PROPDMGEXP == "B", ]$PROPDMG * 10^9

# Convert Crop Damage 
cleanStormData$CROPDMGDOLLARS = 0
cleanStormData[cleanStormData$CROPDMGEXP == "H", ]$CROPDMGDOLLARS = cleanStormData[cleanStormData$CROPDMGEXP == "H", ]$CROPDMG * 10^2
cleanStormData[cleanStormData$CROPDMGEXP == "K", ]$CROPDMGDOLLARS = cleanStormData[cleanStormData$CROPDMGEXP == "K", ]$CROPDMG * 10^3
cleanStormData[cleanStormData$CROPDMGEXP == "M", ]$CROPDMGDOLLARS = cleanStormData[cleanStormData$CROPDMGEXP == "M", ]$CROPDMG * 10^6
cleanStormData[cleanStormData$CROPDMGEXP == "B", ]$CROPDMGDOLLARS = cleanStormData[cleanStormData$CROPDMGEXP == "B", ]$CROPDMG * 10^9

```

####View the headers again to check the changes/addtions to the data
```{r}
head(cleanStormData)
```

####Load the libraries need to produce the grafics 
```{r}
#load libraries
library(ggplot2)
library(gridExtra)
```


####Group fatality data by event type and sum the result
```{r}
fatalities <- aggregate(FATALITIES ~ EVTYPE, data=cleanStormData, sum)
```

####Group the injury data by event type and sum teh result
```{r}
injuries <- aggregate(INJURIES ~ EVTYPE, data = cleanStormData, sum)
```

####Sort the fatality data
```{r}
#sort the fatality data
fatalities <- fatalities[order(-fatalities$FATALITIES), ][1:20, ]

#group by event Type
fatalities$EVTYPE <- factor(fatalities$EVTYPE, levels = fatalities$EVTYPE)
#get header
head(fatalities)
```

####Sort the injury data
```{r}
#Sort the injury data
injuries <- injuries[order(-injuries$INJURIES), ][1:20, ]

#group the injury data
injuries$EVTYPE <- factor(injuries$EVTYPE, levels = injuries$EVTYPE)

#get header
head(injuries)
```

####Set the fatality plot proamters
```{r}
fatalityPlot = ggplot(fatalities, aes(x = EVTYPE, y = FATALITIES, theme_set(theme_bw()))) + 
  geom_bar(stat = "identity", fill = "red") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 4)) + 
  xlab("Event Type") + 
  ylab("Fatalities") + 
  ggtitle("Fatality by Event Types") +
  theme(plot.title = element_text(size = 10))
```

####Set the injury plot pramaters
```{r}
injuryPlot = ggplot(injuries, aes(x = EVTYPE, y = INJURIES, theme_set(theme_bw()))) + 
  geom_bar(stat = "identity", fill = "blue") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 4)) + 
  xlab("Event Type") + 
  ylab("Injuries") + 
  ggtitle("Injury by Event Types") +
  theme(plot.title = element_text(size = 10))
```

####Plot both the fatality and the injury data side by side
```{r}
grid.arrange(fatalityPlot, injuryPlot, ncol = 2, top = "Most Harmful Events By Population Health")
```

####Organize and aggragate the data and group to Event Type and store in object "damage"
```{r}
damage <- aggregate(PROPDMGDOLLARS + CROPDMGDOLLARS ~ EVTYPE, data=cleanStormData, sum)
names(damage) = c("EVENT_TYPE", "TOTAL_DAMAGE")
```

####Get the most damage event in the US 
```{r}
damage <- damage[order(-damage$TOTAL_DAMAGE), ][1:20, ]
damage$EVENT_TYPE <- factor(damage$EVENT_TYPE, levels = damage$EVENT_TYPE)
```

####Get the header of the most damage event
```{r}
head(damage)
```

####Plot the most damage event in the US
```{r}
ggplot(damage, aes(x = EVENT_TYPE, y = TOTAL_DAMAGE, theme_set(theme_bw()))) +
  geom_bar(stat = "identity", fill = "blue") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  xlab("Event Type") + 
  ylab("Total Damage in US Dollor") + 
  ggtitle("the US greatest economic consequences in Property & Crop Damage by top 20 Weather Events")
```
